Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2020-06-11 output/paper/paper.tex
Updates from Overleaf

% Customization
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}

\usepackage{amsmath}
\newcommand{\mv}[1]{{\boldsymbol{\mathrm{#1}}}}

@@

\title{Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden}
\date{\today}
\author{
	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
	Joacim Rocklöv\thanks{Department of Public Health and Clinical Medicine, Section of Sustainable Health, Umeå University}\thanks{Heidelberg Institute of Global Health, University of Heidelberg} \and
	Jonas Wallin\thanks{Department of statistics, Lund university, \href{mailto:jonas.wallin81@gmail.com}{jonas.wallin81@gmail.com}}
}

@@

	\noindent Keywords:

@@

\begin{itemize}
	\item it does not relay any previous trend in the data,
	\item we can generate confidence bounds for the uncertainty about daily true frequencies,
	\item the uncertainty estimates can be carried over to epidemiological models to help create more realistic models.
\end{itemize}

@@

\chapter{Applying the model to COVID-19 in Sweden}
We propose a Bayesian version of the removal model that assumes an overdispersed binomial distribution for the daily observations of deaths in Sweden in COVID-19. We then calculate the posterior distribution, prediction median and 95\% uncertainty intervals of the expected deaths from the reported deaths on each specific day. The method and algorithm is thoroughly described in the Supplementary Information.

@@

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated. SCRPS is a measure of accuracy that rewards precision, similar to a Brier score (see definition in Appendix). The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI´s that cover the true value.

Benchmark and model point estimates are similarly close to the truth. The model produces tighter confidence intervals however, except during the very first day when only a single report is available.

@@

Moreover, deaths are reported from two distinct populations that seem to follow different trends. At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths. But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes. However, wew have noted a clear decline in proportions of deaths reported the two first working days. In early april .\% of deaths where reported within the first two working days wheras in late may only \% where reported during the same time interval. We adress this by assuming that the two deaths from the two first working days comes from a different population then the remainder of days.

@@

\chapter{Conclusion}
In this paper, we provide a simple method to accurately nowcast daily Covid-19 statistics that are reported with delay. By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends. Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during this pandemic.

>>>>

% Customization
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{amsthm}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcommand{\mv}[1]{{\boldsymbol{#1}}}

@@

\title{Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden}
\date{\today}
\author{
	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
	Joacim Rocklöv\thanks{Department of Public Health and Clinical Medicine, Umeå University,\href{mailto:joacim.rocklov@umu.se}{joacim.rocklov@umu.se }}\thanks{Heidelberg Institute of Global Health, University of Heidelberg}  \and
	Jonas Wallin\thanks{Department of statistics, Lund university, \href{mailto:jonas.wallin81@gmail.com}{jonas.wallin81@gmail.com}}
}

@@

	%\noindent Keywords:

@@

\begin{itemize}
	\item it does not relay any previous trend in the data,
	\item we can generate prediction intervals for the uncertainty about daily true frequencies,
	\item the uncertainty estimates can be carried over to epidemiological models to help create more realistic models.
\end{itemize}

@@

\chapter{Applying the model to COVID-19 in Sweden}
We propose a Bayesian version of the removal model that assumes an overdispersed binomial distribution for the daily observations of deaths in Sweden in COVID-19. We then calculate the posterior distribution, prediction median and 95\% prediction intervals of the expected deaths from the reported deaths on each specific day. The method and algorithm is thoroughly described in the Supplementary Information.

@@

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated. SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \cite{bolin2019scale}. The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI´s that cover the true value.

Benchmark and model point estimates are similarly close to the truth. The model produces tighter prediction intervals however, they are too tight between day 8-5 lags to reporting, see Figure \ref{fig:model_metrics}, this is likely due to that Swedish Public Health Agency does a database search during only certain days to match people reported dead to Covid infected. This type of reporting is currently not implemented in the model hence the poorer performance during these lags.

@@

Moreover, deaths are reported from two distinct populations that seem to follow different trends. At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths. But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes. However, we have noted a clear decline in proportions of deaths reported the two first working days. For example the number of deaths occurring at the second of April $\approx 30\%$ of deaths where reported within the first two working days whereas for the eighteens of May only $\approx 10\%$ where reported during the two first working days. We address this by assuming that the deaths reported during the two first working days comes from a different population then the remainder of days.

@@

\chapter{Conclusion}
In this paper, we provide a method to accurately nowcast daily Covid-19 statistics that are reported with delay. By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends. Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during this pandemic.
